---
title: "Example: Probit Model"
description: >
  See the {ino} package in action for the probit likelihood optimization.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Example: Probit Model}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ref.bib
link-citations: true
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "figures/probit-",
  fig.dim = c(8, 6), 
  out.width = "75%"
)
library("ino")
options("ino_verbose" = TRUE) 
set.seed(1)
ggplot2::theme_set(ggplot2::theme_minimal())
```

The probit model is a widely employed statistical tool for analyzing discrete choice behavior in various fields, including transportation [@Bolduc:1999; @Shin:2015] and marketing [@Allenby:1998; @Haaijer:1998; @Paap:2000]. The estimation of probit model parameters typically involves the numerical maximization of the likelihood function. However, this approach can be computationally expensive and may encounter problems in reaching the global optimum, especially when dealing with complex models. In this vignette, we utilize the `{ino}` package to investigate the influence of initialization on the numerical maximization of the probit likelihood function.

## Model formulation

In our model formulation, we consider a scenario where a total of $N$ deciders are faced with choosing among $J \geq 2$ alternatives at each of the $T$ choice occasions. The choice made by decider $n$ at occasion $t$ is denoted as $y_{nt}$ and can take values in the set $\{1, \dots, J\}$. For more comprehensive details on the probit model and its estimation, please refer to the works by @Train:2009 and @Bhat:2011.

We assume that the choices are rational, meaning that the selected alternative $y_{nt}$ corresponds to the alternative with the highest utility among the available options. The utility for decider $n$ at occasion $t$ is represented by a vector $U_{nt} \in \mathbb{R}^J$, where each entry of the vector corresponds to the utility associated with a specific alternative. The probit model explains the utility vector as $$U_{nt} = X_{nt} b + \epsilon_{nt},$$ where $X_{nt}$ is a $J\times P$ matrix containing $P$ characteristics for each alternative, $b$ is a coefficient vector of length $P$, and $\epsilon_{nt} \sim N(0,\Sigma)$ denotes the vector of jointly normally distributed errors, which capture unobserved influences on the utility.

The probit model (like any utility model) is invariant to the level and scale of the utilities $U_{nt}$. We ensure identifiability by considering utility differences, which reduces $\Sigma$ from $J$ to $J-1$ dimensions, and fixing the first entry of $b$ to $1$.

To account for preference heterogeneity across decision-makers, the mixed probit model incorporates decider-specific coefficient vectors as $\beta_n \sim N(b, \Omega)$. In the degenerate case where $\Omega = 0$, all decision-makers share the same preferences, and $\beta_n \equiv b$.

The goal of the researcher is to estimate the values for $b$, $\Omega$, and $\Sigma$ based on a set of observed choice data. The most common approach for estimation is the maximum likelihood method. Let $\theta$ represent the vector of identified parameters, which includes $P-1$ coefficients of $b$, $P(P+1)/2$ coefficients of $\Omega$, and $J(J-1)/2$ coefficients of the differenced matrix $\Sigma$. To ensure that the estimates result in proper covariance matrices, the optimization is performed over the Cholesky factors. It is important to note that the length of the parameter vector $\theta$ increases quadratically with both the number of alternatives $J$ and the number of choice characteristics $P$, indicating that numerical optimization becomes computationally demanding for complex models with a high number of alternatives or choice covariates.

The maximum likelihood estimate $\hat{\theta}$ is obtained by solving $$\hat{\theta} = \arg \max_\theta \log \sum_{n,t,j} 1(y_{nt} = j) \int 1(j = \arg \max U_{nt}) \phi(\epsilon_{nt}) d \epsilon_{nt},$$ where $1(\cdot)$ denotes the indicator function and $\phi(\cdot)$ represents the Gaussian density. The integral part of the equation does not have a closed-form expression, thus requiring numerical approximation. Here, we utilize the `mvtnorm::GenzBretz()` algorithm by @Genz:2009 for this purpose.

## Data simulation and likelihood computation

The `{ino}` package provides the `sim_mnp()` function, which enables simulation of choice data from a probit model. Prior to utilizing this function, it is necessary to define the function `X(n, t)`, which generates a matrix representing the choice characteristics for decision maker $n$ at choice occasion $t$. The matrix should have dimensions of $J \times P$, where $J$ corresponds to the number of available alternatives, and $P$ indicates the number of characteristics describing each alternative.

For our simulation, we have a choice setting with $J = 3$ alternatives, characterized by $P = 2$ attributes. The values in the first column of $X_{nt}$ are drawn from a normal distribution $\mathcal{N}(\mu = 10, \sigma = 3)$, while the values in the second column are drawn from $\mathcal{N}(\mu = 0, \sigma = 0.3)$. This design reflects the common situation of observing choice covariates on different scales.

```{r, choice covariates, eval = TRUE}
X <- function(n, t) {
  J <- 3
  cbind(stats::rnorm(J, mean = 10, sd = 3), stats::rnorm(J, mean = 0, sd = 0.3))
}
X(n = 1, t = 1)
```

We simulate choice data for $N = 100$ deciders at $T = 20$ choice occasions from the probit model defined by the parameter values $b = \begin{pmatrix} 1 & -10 \end{pmatrix}^\top$, $\Omega = \begin{pmatrix} 0.2 & 0.5 \\ 0.5 & 2 \end{pmatrix}$, and $\Sigma = \begin{pmatrix} 1 & -0.5 & 0.2 \\ -0.5 & 1 & 0.2 \\ 0.2 & 0.2 & 1 \end{pmatrix}$:

```{r, simulate data, eval = TRUE}
N <- 100
Tp <- 20
b <- c(1, -10)
Omega <- matrix(c(0.2, 0.5, 0.5, 2), 2, 2)
Sigma <- matrix(c(1, -0.5, 0.2, -0.5, 1, 0.2, 0.2, 0.2, 1), 3, 3)
probit_data <- sim_mnp(N, Tp, J = 3, P = 2, b, Omega, Sigma, X, seed = 1)
```

The `probit_data` object is a `data.frame` with the decider index `n`, the choice occasion index `t`, the choice `y`, and the choice characteristics `X`:

```{r, head of data, eval = TRUE}
head(probit_data)
```

The `probit_data` object includes the attribute `true`, which contains the true and *identified* parameter values. These values consist of the mean effects $b$, excluding the first element, the elements $o$ of the lower-triangular Cholesky root of $\Omega$, and the elements $l$ of the lower-triangular Cholesky root of the differenced covariance matrix $\Sigma$ (with respect to alternative $J$):

```{r, true parameter, eval = TRUE}
round(theta <- attr(probit_data, "true"), 2)
```

The probit likelihood function is implemented as `f_ll_mnp()` and can be evaluated by providing a parameter vector `theta` and a data set `probit_data`:

```{r, likelihood evaluation, eval = TRUE}
f_ll_mnp(theta = theta, data = probit_data)
```

## Setup

To analyze the outcome of the numerical likelihood optimization for the multinomial mixed probit model, we apply the `{ino}` package:

- We first define a `Nop` object by setting the target function `f = f_ll_mnp`, the number of parameters to `npar = 7`, and `data = probit_data`. 
- Then, we apply the `stats::nlm` optimizer with a limit of 1000 iterations. Since this optimizer minimizes the objective function instead of maximizing it, we set `neg = TRUE` to compute the negative log-likelihood value.

```{r, define Nop, eval = FALSE}
probit_ino <- Nop$new(f = f_ll_mnp, npar = 7, data = probit_data, neg = TRUE)$
  set_optimizer(optimizer_nlm(iterlim = 1000))
```

The true parameter vector `theta` is saved to assess the convergence of the optimization runs to the global optimum:

```{r, set true parameter, eval = FALSE}
probit_ino$true_parameter <- theta
```

The initial `Nop` object looks as follows:

```{r, print initial Nop object, eval = TRUE}
print(probit_ino)
```

## Random initialization

As a benchmark, we optimize `runs = 100` times using random initial values drawn from a standard normal distribution:

```{r, optimize with random initial values, eval = FALSE}
probit_ino$optimize(initial = "random", runs = 100, label = "random")
```

## Initializing using estimates from a data subsample

Next, we obtain starting values by estimating a probit model on a subset of the data. In this example, we randomly select 20\% of the data points for the subset. Our aim is to initialize the full model close to its global optimum by investing a small computational effort in estimating the reduced model:

```{r, optimize on subset, eval = FALSE}
probit_ino$
  reduce("data", how = "random", proportion = 0.2)$
  optimize(initial = "random", runs = 100, label = "subset")$
  reset_argument("data")$
  continue()
```

The `$reduce()` method offers additional options for selecting the composition of the subset, which can further enhance the initialization process:

- `how = "first"` and `how = "last"`: These options allow to select the subset from the beginning or the end of the data, respectively, focusing on specific segments of the data set.

- `how = "similar"` and `how = "dissimilar"`: These options utilize k-means clustering to identify subsets of similar or dissimilar data points. By clustering the data based on similarity, a subset that represents a distinct group within the data set can be selected. This approach can be particularly useful when there are distinct patterns or characteristics within the data.

```{r, deviation, eval = TRUE, warning = FALSE}
probit_ino$deviation(
  reference = probit_ino$true_parameter, which_run = c("random", "subset"),
  parameter_labels = c("b.2", "o.11", "o.21", "o.22", "l.11", "l.21", "l.22"),
  ylim = c(-10, 10)
)
```
	
## Estimation with standardized covariates

In choice settings, covariates can have varying scales, which can result in differences in the magnitudes of model parameters. To enhance numerical stability during estimation, a common strategy is to standardize the covariates prior to the estimation. Here, we evaluate the effectiveness of standardizing covariates using the `$standardize()` method. By setting `ignore = 1:3`, we exclude the first columns containing indices from the standardization.

```{r, optimize with standardized data, eval = FALSE}
probit_ino$
  standardize("data", by_column = TRUE, ignore = 1:3)$
  optimize(initial = "random", runs = 100, label = "standardized")$
  reset_argument("data")
```

We also combine standardization and subsampling:

```{r, optimize with standardize and subset, eval = FALSE}
probit_ino$
  standardize("data", by_column = TRUE, ignore = 1:3)$
  reduce("data", how = "random", proportion = 0.2)$
  optimize(initial = "random", runs = 100, label = "standardized_subset")$
  reset_argument("data")$
  standardize("data", by_column = TRUE, ignore = 1:3)$
  continue()
```

## Results

Both the subset and standardize approach improve the optimization time significantly in comparison to the random initialization. The combination of both strategies is twice as fast on average.

```{r, times, eval = TRUE, message = FALSE}
plot(probit_ino, by = "label", relative = TRUE, xlim = c(-1, 3))
```
	
```{r, summary, eval = TRUE, warning = FALSE, message = FALSE}
library("dplyr")
summary(probit_ino, which_element = c("seconds", "label")) %>% 
  group_by(label) %>% 
  summarize(
    mean_seconds = mean(seconds, na.rm = TRUE),
    sd_seconds = sd(seconds, na.rm = TRUE)
  ) %>%
  arrange(mean_seconds)
```
	
## References

