---
title: "Example: Hidden Markov Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Example: Hidden Markov Model}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ref.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.dim = c(8, 6), 
  out.width = "75%"
)
# library("ino")
devtools::load_all() # remove later
set.seed(1)
hmm_ino <- ino::hmm_ino
```

This vignette^[The vignette was build using R `r paste(R.Version()[c("major","minor")], collapse = ".")` with the {ino} `r utils::packageVersion("ino")` package.] describes the workflow of the {ino} package for the likelihood optimization of an hidden Markov model (HMM). For more technical details about HMMs see, e.g., @zucchini2016.

## Data

The example data set considered throughout this vignette covers a time series of share returns from the Deutsche Bank stock. The data set is freely accessible via Yahoo Finance and can be downloaded via the {fHMM} package [@Oelschläger:2023b]: 

```{r, message = FALSE, warning = FALSE}
library("dplyr")
data <- fHMM::download_data(symbol = "DBK.DE", from = "2000-01-01")
db_data <- data %>%
  as_tibble() %>%
  reframe(
    date = as.Date(Date, format = "%Y-%m-%d"),
    logreturn = c(NA, diff(log(Close), lag = 1))
  ) %>%
  filter(!is.na(logreturn)) %>%
  print()
```

The time series looks as follows:

```{r, message = FALSE, warning = FALSE}
library("ggplot2")
ggplot(db_data, aes(x = date, y = logreturn)) +
  geom_point() +
  geom_line() +
  scale_x_date() +
  scale_y_continuous(labels = scales::label_percent())
```

As the share returns are continuous and can take both negative and positive values, we consider an HMM with Gaussian state-dependent distributions.

## Model fitting

We consider a 2-state (`N = 2`) Gaussian-HMM with six parameters (`npar = 6`) to be estimated:

- two for the transition probability matrix (t.p.m.)
- two for the means of the state-dependent distributions
- two for the standard deviations of the state-dependent distributions

The likelihood for a Gaussian-HMM is given by the function `f_ll_hmm()` provided by the {ino} package. The argument `neg =  TRUE` indicates that we minimize the negative log-likelihood, and `set_optimizer(optimizer_nlm())` selects the optimizer `nlm()` (see the introductory vignette).

```{r, eval = FALSE}
hmm_ino <- Nop$new(
    f = f_ll_hmm, 
    npar = 6, 
    data = db_data$logreturn, 
    N = 2, 
    neg = TRUE
  )$
  set_optimizer(optimizer_nlm())
```

### Random initialization

We first use randomly chosen starting values. As the first two starting values belong to the t.p.m., we draw starting values from a $\mathcal{U}(-2,-1)$ distribution --- as we use the multinomial logit link to ensure that the probabilities are between 0 and 1, a value of -1.5 correspond to probabilities of staying in state 1 and 2 of about 0.81. For the two means, we draw two random numbers from the standard normal distribution, as the time series above indicates that the most of the returns are fairly close to zero. The starting values for the standard deviations are drawn from a $\mathcal{U}(0.5,2)$ distribution (note that we exponentiate the standard deviations in the likelihood as they are constrained to be positive, and hence we log-transform the starting values).

The `$optimize()` method with argument `initial = sampler` then performs `runs = 50` optimization runs with starting values drawn from the selected distributions. For comparisons with other optimization strategies (which will be carried out below), it's useful to set a label.

```{r, eval = FALSE}
sampler <- function() {
  c(stats::runif(2, -2, -1), stats::rnorm(2), log(stats::runif(2, 0.5, 2)))
}
hmm_ino$optimize(initial = sampler, runs = 50, label = "random")
```

### Educated guesses

For selecting fixed starting values, we consider three sets of starting values that fall in the ranges considered above. These can be considered as "educated guesses" and are likely to be close to the global optimum:

```{r, eval = FALSE}
tpm_entry_1 <- tpm_entry_2 <- c(-2, -2.5)
mu_1 <- c(0, -0.05)
mu_2 <- c(0, 0.05)
sd_1 <- c(log(0.1), log(0.5))
sd_2 <- c(log(0.75), log(1))

starting_values <- asplit(expand.grid(
  tpm_entry_1, tpm_entry_2, mu_1, mu_2, sd_1, sd_2), MARGIN = 1)

hmm_ino$optimize(initial = starting_values, label = "educated_guess")
```

### Subset initialization

The time series of share returns consists of `r nrow(db_data)` observations, so it 
To illustrate the subset initialization strategy, we consider the first 20\% of the observation using `$reduce()`. The starting values for this subset are chosen randomly from the same distributions `sampler` as considered above. We again use the function `$optimize()` to fit the HMM, but now to the subset containing the first 20\% of the observations. With `$continue()`, we then use the estimates obtained from the subsets as initial values to fit the model to the entire data set.

```{r, eval = FALSE}
hmm_ino$
  reduce("data", how = "first", prop = 0.2)$
  optimize(initial = sampler, runs = 50, label = "subset")$
  continue()$
  reset_argument("data")
```

As a final step in the previous code snippet, we use `reset_argument()` to obtain our actual data set consisting of *all* observations. If we were to skip this step, all future optimization runs would be made on the subset.

## Evaluating the optimization runs

### Local optima

Selecting the starting values for HMMs is a well-known issue, as poor starting values may likely result in local optima. Other R packages designed to fit HMMs discuss this topic in more detail (see, e.g., https://cran.r-project.org/package=moveHMM). We thus first evaluate the optimizations by comparing the likelihood values at convergence, which can be displayed using `optima()`:

```{r eval=TRUE}
hmm_ino$optima()
```

The frequency table indicates that `r hmm_ino$optima() %>% arrange(value) %>% slice(1) %>% pull(frequency)` out of `r sum(hmm_ino$optima()$frequency)` runs converged to the smallest negative likelihood value, which appears to be the global optimum (note these are the negative log-likelihood values). However, in `r hmm_ino$optima() %>% arrange(value) %>% slice(2:nrow(.)) %>% pull(frequency)` runs we gut stuck in a local optimum.

Using `summary()`, we can investigate the computation time ("seconds"), the resulting optimum ("value") and the corresponding initialization strategy ("label") of all runs (here, only the first 10 are shown): 

```{r eval=TRUE}
hmm_ino$summary(c("value", "seconds", "label")) %>% 
  head(n = 10)
```


The final parameter estimates, i.e., the parameters associated with the global optimum, are stored as `best_parameter`:

```{r eval=TRUE}
hmm_ino$best_parameter
```

and the corresponding likelihood value can be obtained as

```{r eval=TRUE}
hmm_ino$best_value
```


### Optimization time

We use the `plot()` function to investigate the computation time. Intuitively, optimization runs based on our educated guesses should be faster than runs with random starting values, since we have carefully chosen these starting values by investigating the data:

```{r eval=TRUE}
plot(hmm_ino, by = "label")
```

Here, relative computation times are shown --- setting `relative = FALSE` in `plot()` gives the absolute times. Using the output provided by `summary()` and some data manipulation functions provided by the package {dplyr}, we can also compute the average time per strategy:

```{r warning = FALSE, message = FALSE, eval=TRUE}
summary(hmm_ino, c("label", "seconds")) %>% 
  group_by(label) %>% 
  summarise(avg_time = mean(seconds, na.rm = TRUE))
```

This comparison can however be considered somehow 'unfair', as not all optimization runs lead to the global optimum --- optimization runs that lead to local optima may then have less iterations and hence lower computation time. We can first check which runs lead to the apparent global optimum of `r hmm_ino$best_value`. For that, we will again use functions from {dplyr}.

```{r eval=TRUE}
summary(hmm_ino, c("label", "value", "seconds")) %>% 
  mutate(global_optimum = ifelse(value < -14157, 1, 0)) %>% 
  group_by(label) %>% 
  summarise(proportion_global_optimum = mean(global_optimum, na.rm = TRUE))
```

While for the educated guesses, most of the runs converge to the global optimum, the random initialization and subset initialization strategy with 5% of the data get stuck more often in local maxima.

Let's again compare the average computation time, but only for those runs that lead to the global optimum:

```{r eval=TRUE}
summary(hmm_ino, c("label", "value", "seconds")) %>% 
  filter(value < -14157) %>% 
  group_by(label) %>% 
  summarise(mean_time = mean(seconds))
```

For those runs that converged to the global optimum, we can again compare the computation time via boxplots:

```{r eval=TRUE}
summary(hmm_ino, c("label", "value", "seconds")) %>% 
  filter(value < -14157) %>% 
  ggplot(aes(x = "", y = seconds)) +
    scale_y_continuous() +
    geom_boxplot() +
    facet_wrap("label", labeller = "label_both", nrow = 1) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank()
    ) +
    ylab("seconds")
```

While computation is fastest for the educated guesses, subset initialization appears to be a promising approach.

## References
