---
title: "Workflow for using ino"
author: "Lennart Oelschläger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow for using ino}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.dim = c(8, 6),
  out.width = "100%"
)
```

This vignette describes the workflow for using ino.

### 0. Load ino

```{r, setup}
# install.packages("ino")
library(ino)
library(magrittr) # Used for better code readability.
library(ggplot2)
```

### 1. Introduce your function to ino

Say we want to numerically optimize a version of the [Himmelblau's function](https://en.wikipedia.org/wiki/Himmelblau%27s_function).

```{r, define function}
himmelblau <- function(x, c, d) (x[1]^2 + x[2] + c)^2 + (x[1] + x[2]^2 + d)^2
```

For `c <- -11` and `y <- -7`, the function is known to have one local maximum and four identical local minima, so it is worth putting a thought on the initialization.....

```{r, visualize function, fig.cap = "Contour plot of the Himmelblau's function"}
x <- y <- seq(-5, 5, 0.1)
#z <- outer(x, y, Vectorize(function(x, y) himmelblau(c(x, y), c = -11, d = -7)))
#contour(x, y, z)
data_plot <- expand.grid(x, y)
data_plot$z <- purrr::map2_dbl(data_plot$Var1, data_plot$Var2, function(x, y) himmelblau(c(x, y), c = -11, d = -7))
data_plot %>% 
  ggplot(aes(x = Var1, y = Var2, z = z)) + geom_contour_filled() +
  theme_minimal() + xlab("x") + ylab("y") +
  theme(legend.title = element_blank())
```
<!-- TODO: Add better visualization. Perhaps own visualization function? -->

We introduce the function to ino via:

```{r, introduce f}
analysis <- himmelblau %>% set_f(npar = 2, c = -11, d = -7)
```

This does three things:

1. We created `analysis` which is an object of class `ino`. Insights about the initialization will be saved here.

2. We specified that we optimize over 2 parameters.

3. We fixed the parameters `c` and `d`.

> Remark: If your function is a likelihood function, then this is the right time to also introduce your dataset to ino. See the vignette ... on how to do this.

### 2. Select a numerical optimizer

Next, we need to specify a numerical optimizer, for example nlm from the stats package. We can also set parameters of the optimizer, for example the maximum number of iterations `iterlim`.

```{r, set optimizer}
analysis %<>% set_optimizer(optimizer = stats::nlm, iterlim = 1000)
```

> Remark: You can specify a list of optimizers here to compare them. See the vignette ... for details.

Now, all is setup to try different initialization strategies.

### 3. Execute initialization strategies

There is a separate function for each initialization strategy in ino. Let's try random initialization:

```{r, random initialization}
analysis %<>% random_initialization(runs = 100, lower = -10, upper = 10, sampler = "normal")
```

We initialized `runs = 100` times at random samples from a normal distributions restricted on the interval `lower = -10` to `upper = 10`.

Let's also try fixed initialization at the origin:
```{r, fixed initialization}
analysis %<>% fixed_initialization(at = c(0,0))
```

> Remark: Each initialization strategy implemented in ino is documented in its own vignette.

### 4. Evaluate the results

The package provides several functions to evaluate the initialization strategies, for example:

```{r, evaluate optima}
analysis %>% optima
```

```{r, evaluate time}
analysis %>% optimization_time
```

> Remark: Each evaluation function again is documented in its own vignette.
