---
title: "Introduction"
description: >
  Learn how to get started with the {ino} package.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ref.bib
link-citations: true
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "figures/ino-",
  fig.dim = c(8, 6), 
  out.width = "75%",
  # all optimizations are pre-computed to save building time
  eval = FALSE
)
library("ino")
options("ino_verbose" = TRUE)
data("mixture_ino")
set.seed(1)
ggplot2::theme_set(ggplot2::theme_minimal())
```

## Motivation

Optimization is of great relevance in many fields, including finance (portfolio optimization), engineering (minimizing air resistance), and statistics (likelihood maximization for model fitting). Often, the optimization problem at hand cannot be solved analytically, for example when explicit formulas for gradient or Hessian are not available. In these cases, numerical optimization algorithms are helpful. They iteratively explore the parameter space, guaranteeing to improve the function value over each iteration, and eventually converge to a point where no more improvements can be made [@Bonnans:2006]. In R, several functions are available that can be applied to numerically solve optimization problems, including (quasi) Newton (`stats::nlm()`, `stats::nlminb()`, `stats::optim()`), direct search (`pracma::nelder_mead()`), and conjugate gradient methods (`Rcgmin::Rcgmin()`). The [CRAN Task View: Optimization and Mathematical Programming](https://CRAN.R-project.org/view=Optimization) provides a comprehensive list of packages for solving optimization problems. 

One thing that all of these numerical optimizers have in common is that initial parameter values must be specified, i.e., the point from where the optimization is started. Optimization theory [@Nocedal:2006] states that the choice of an initial point has a large influence on the optimization result, in particular convergence time and rate. In general, starting in areas of function saturation increases computation time, starting in areas of non-concavity leads to convergence problems or convergence to local rather than global optima. Consequently, numerical optimization can be facilitated by 

1. analyzing the initialization effect for the optimization problem at hand and

2. putting effort on identifying good starting values.

However, it is generally unclear what good initial values are and how they might affect the optimization. Therefore, the purpose of the `{ino}` R package^[*ino* is an acronym for ***i***nitialization of ***n***umerical ***o***ptimization.] is to provide a comprehensive toolbox for 

1. evaluating the effect of the initial values on the optimization,

2. comparing different initialization strategies,

3. and comparing different optimizers.

## Package functionality

To specify an optimization problem in `{ino}`, we use an object-oriented framework based on the `{R6}` package [@Chang:2021]. The general workflow is to first create
a `Nop` object^[*Nop* is an acronym for ***n***umerical ***o***ptimization ***p***roblem.], and then apply methods to change the attributes of that object, e.g., to optimize the function and investigate the optimization results:

- The starting point for working with `{ino}` is to initialize a `Nop` object via `object <- Nop$new()`.

- Next, use the method `$set_optimizer()` to define one or more numerical optimizer.

- Then, `$evaluate()` evaluates and `$optimize()` optimizes the objective function.

- For analyzing the results, `$optima()` provides an overview of all identified optima, and the `$plot()` and `$summary()` methods summarize the optimization runs.

- The methods `$standardize()` and `$reduce()` are available to advantageously transform the optimization problem.

We illustrate these methods in the following application.

## Workflow

We demonstrate the basic `{ino}` workflow in the context of likelihood maximization, where we fit a two-class Gaussian mixture model to Geyser eruption times from the popular `faithful` data set that is provided via base R.

> **Remark:** Optimization in this example is very fast. This is because the data set is relatively small and we consider a model with two classes only. Therefore, it might not seem relevant to be concerned about initialization here. However, the problem scales: optimization time will rise with more data and more parameters, in which case initialization becomes a greater issue, see for example @Shireman:2017. Additionally, we will see that even this simple optimization problem suffers heavily from local optima, depending on the choice of initial values.

### The optimization problem

The `faithful` data set contains information about eruption times (`eruptions`) of the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.

```{r, faithful data, eval = TRUE}
str(faithful)
```

The data histogram hints at two clusters with short and long eruption times, respectively.

```{r, faithful, warning = FALSE, eval = TRUE}
library("ggplot2")
ggplot(faithful, aes(x = eruptions)) + 
  geom_histogram(aes(y = after_stat(density)), bins = 30) + 
  xlab("eruption time (min)") 
```

For both clusters, we assume a normal distribution, such that we consider a mixture of two Gaussian densities for modeling the overall eruption times. The log-likelihood function is defined by

\begin{equation}
\ell(\boldsymbol{\theta}) = \sum_{i=1}^n \log\Big( \lambda \phi_{\mu_1, \sigma_1^2}(x_i) + (1-\lambda)\phi_{\mu_2,\sigma_2^2} (x_i) \Big),
\end{equation}

where the sum goes over all observations $1, \dots, n = 272$, $\phi_{\mu_1, \sigma_1^2}$ and $\phi_{\mu_2, \sigma_2^2}$ denote the normal density for the first and second cluster, respectively, and $\lambda$ is the mixing proportion. The parameter vector to be estimated is thus $\boldsymbol{\theta} = (\mu_1, \mu_2, \sigma_1, \sigma_2, \lambda)$. As there exists no closed-form solution for the maximum likelihood estimator $\boldsymbol{\theta}^* = \arg\max_{\boldsymbol{\theta}} \ell(\boldsymbol{\theta})$, we need numerical optimization for finding the function optimum. 

The following function calculates the log-likelihood value given the parameter vector `theta` and the observation vector `data`. 

> **Remark:** We restrict the standard deviations `sd` to be positive (via the exponential transformation) and `lambda` to be between 0 and 1 (via the logit transformation). The function returns the negative log-likelihood value by default (`neg = TRUE`). This is necessary because most R optimizers only minimize (e.g., `stats::nlm`), where we can use the fact that $\arg\max_{\boldsymbol{\theta}} \ell(\boldsymbol{\theta}) = \arg\min_{\boldsymbol{\theta}} -\ell(\boldsymbol{\theta})$.

```{r, mixture ll, eval = TRUE}
normal_mixture_llk <- function(theta, data, neg = TRUE){
  stopifnot(length(theta) == 5)
  mu <- theta[1:2]
  sd <- exp(theta[3:4])
  lambda <- plogis(theta[5])
  llk <- sum(log(
    lambda * dnorm(data, mu[1], sd[1]) + 
      (1 - lambda) * dnorm(data, mu[2], sd[2])
    ))
  ifelse(neg, -llk, llk)
}
normal_mixture_llk(theta = 1:5, data = faithful$eruptions)
```

### Another optimization approach: the expectation-maximization (EM) algorithm

Solving $\boldsymbol{\theta}^* = \arg\max_{\boldsymbol{\theta}} \ell(\boldsymbol{\theta})$ requires numerical aid because  $\frac{\text{d}}{\text{d}\boldsymbol{\theta}} \ell(\boldsymbol{\theta})$ does not have a closed-form solution. However, if we knew the class membership of each observation, the optimization problem would collapse to independent maximum likelihood estimation of two Gaussian distributions, which then can be solved analytically. This observation motivates the so-called expectation-maximization (EM) algorithm [@Dempster:1977], which iterates through the following steps:

1. Initialize $\boldsymbol{\theta}$ and compute $\ell(\boldsymbol{\theta})$.
2. Calculate the posterior probabilities for each observation's class membership, conditional on $\boldsymbol{\theta}$.
3. Calculate the maximum likelihood estimate $\boldsymbol{\bar{\theta}}$ conditional on the posterior probabilities from step 2.
4. Evaluate $\ell(\boldsymbol{\bar{\theta}})$. Now stop if the likelihood improvement $\ell(\boldsymbol{\bar{\theta}}) - \ell(\boldsymbol{\theta})$ is smaller than some threshold `epsilon` or some iteration limit `iterlim` is reached. Otherwise, go back to step 2.

The following function implements this algorithm, which we will compare to standard numerical optimization below.

```{r, em algorithm, eval = TRUE}
em <- function(normal_mixture_llk, theta, epsilon = 1e-08, iterlim = 1000, data) {
  llk <- normal_mixture_llk(theta, data, neg = FALSE)
  mu <- theta[1:2]
  sd <- exp(theta[3:4])
  lambda <- plogis(theta[5])
  for (i in 1:iterlim) {
    class_1 <- lambda * dnorm(data, mu[1], sd[1])
    class_2 <- (1 - lambda) * dnorm(data, mu[2], sd[2])
    posterior <- class_1 / (class_1 + class_2)
    lambda <- mean(posterior)
    mu[1] <- mean(posterior * data) / lambda
    mu[2] <- (mean(data) - lambda * mu[1]) / (1 - lambda)
    sd[1] <- sqrt(mean(posterior * (data - mu[1])^2) / lambda)
    sd[2] <- sqrt(mean((1 - posterior) * (data - mu[2])^2) / (1 - lambda))
    llk_old <- llk
    theta <- c(mu, log(sd), qlogis(lambda))
    llk <- normal_mixture_llk(theta, data, neg = FALSE)
    if (is.na(llk)) stop("fail")
    if (llk - llk_old < epsilon) break
  }
  list("neg_llk" = -llk, "estimate" = theta, "iterations" = i)
}
```

### Setup

The optimization problem is specified as a `Nop` object called `mixture_ino`, where

- `f` is the function to be optimized (here `normal_mixture_llk`), 
- `npar` specifies the length of the parameter vector over which `f` is optimized (five in this case),
- and `data` gives the observation vector as required by our likelihood function. 

```{r, initialize mixture_ino}
mixture_ino <- Nop$new(
  f = normal_mixture_llk, 
  npar = 5, 
  data = faithful$eruptions
)
```

The next step concerns specifying the numerical optimizer via the `$set_optimizer()` method. 

> **Remark:** Numerical optimizers must be specified through the unified framework provided by the [{optimizeR}](https://CRAN.R-project.org/package=optimizeR) package [@Oelschläger:2023]. This is necessary because there is no a priori consistency across optimization functions in R with regard to their function inputs and outputs. This would make it impossible to allow for arbitrary optimizers and to compare their results, see [the {optimizeR} README file](https://github.com/loelschlaeger/optimizeR#readme) for details.

It is possible to define any numerical optimizer implemented in R through the {optimizeR} framework. Here, we select two of the most popular ones, `stats::nlm()` and `stats::optim()`:

```{r, mixture_ino optimizer}
mixture_ino$
  set_optimizer(optimizer_nlm(), label = "nlm")$
  set_optimizer(optimizer_optim(), label = "optim")
```

> **Remark:** The previous code chunk makes use of a technique called "method chaining" [see @Wickham:2019, ch. 14.2.1]. This means that `mixture_ino$set_optimizer()` returns the modified `mixture_ino` object, for which we can specify a second optimizer by calling `$set_optimizer()` again. 

We also want to apply the EM algorithm introduced above:

```{r, set em algorithm}
em_optimizer <- optimizeR::define_optimizer(
  .optimizer = em, .objective = "normal_mixture_llk",
  .initial = "theta", .value = "neg_llk", .parameter = "estimate",
  .direction = "min"
)
mixture_ino$set_optimizer(em_optimizer, label = "em")
```

Finally, we can validate our specification:

```{r, validate mixture_ino, eval = FALSE}
mixture_ino$test(verbose = FALSE)
```

### Function evaluation

Once the `Nop` object is specified, evaluating `normal_mixture_llk` at some value for the parameter vector `theta` is simple with the `$evaluate()` method, for example:

```{r, example evaluation, eval = TRUE}
mixture_ino$evaluate(at = 1:5)
```

### Function optimization

Optimization of `normal_mixture_llk` is possible with the `$optimize()` method, for example:

```{r, example optimization, eval = TRUE}
mixture_ino$optimize(
  initial = "random", which_optimizer = "nlm", save_result = FALSE, return_result = TRUE
)
```

The method arguments are:

- `initial = "random"` for random starting values drawn from a standard normal distribution,

- `which_optimizer = "nlm"` for optimization with the above specified `stats::nlm` optimizer,

- `save_result = FALSE` to not save the optimization result inside the `mixture_ino` object (see below),

- and `return_results = TRUE` to directly return the optimization result instead.

The return value is a `list` of:

- `value`, the optimum function value,

- `parameter`, the parameter vector where `value` is obtained,

- `seconds`, the estimation time in seconds,

- `initial`, the starting parameter vector for the optimization,

- and `gradient`, `code`, and `iterations`, which are outputs specific to the `stats::nlm` optimizer.

### Initialization effect

We are interested in the effect of the starting values on the optimization, i.e., whether different initial values lead to different results. We therefore optimize the likelihood function `runs = 100` times at different random starting points (`initial = "random"`) and compare the identified optima:

```{r, random initialization}
mixture_ino$optimize(
  initial = "random", runs = 100, label = "random", save_results = TRUE, seed = 1
)
```

Note:

1. We label the optimization results with `label = "random"`, which will be useful later for comparisons. 
2. We set `save_results = TRUE` to save the optimization results inside the `mixture_ino` object (so that we can use the `$optima()`, `$summary()`, and `$plot()` methods for comparisons, see below). 
3. The `seed = 1` argument ensures reproducibility.

The `$optima()` method provides an overview of the identified optima. Here, we ignore any decimal places by setting `digits = 0` and sort by the optimum function `value`:

```{r, show optima, eval = TRUE}
mixture_ino$optima(digits = 0, which_run = "random", sort_by = "value")
```

The 100 optimization runs with 3 optimizers using random starting values led to `r nrow(na.omit(mixture_ino$optima(digits = 0, which_run = "random")))` different optima (minima in this case, because we minimized `normal_mixture_llk()`), while `r mixture_ino$optima(digits = 0, which_run = "random")[which(is.na(mixture_ino$optima(digits = 0, which_run = "random")$value)), "frequency"]` optimization runs failed. We therefore can already deduce that the initial values have a huge impact on the optimization result.

Looking at this overview optimizer-wise reveals that the `stats::optim` optimizer seems to be most vulnerable to local optima:

```{r, show optima optimizer-wise, eval = TRUE}
mixture_ino$optima(digits = 0, which_run = "random", sort_by = "value", which_optimizer = "nlm")
mixture_ino$optima(digits = 0, which_run = "random", sort_by = "value", which_optimizer = "optim")
mixture_ino$optima(digits = 0, which_run = "random", sort_by = "value", which_optimizer = "em")
```

The two most occurring optima are 421 and 276 with total frequencies of `r mixture_ino$optima(digits = 0, which_run = "random")[which(as.numeric(as.character(mixture_ino$optima(digits = 0, which_run = "random")$value)) == 421), "frequency"]` and `r mixture_ino$optima(digits = 0, which_run = "random")[which(as.numeric(as.character(mixture_ino$optima(digits = 0, which_run = "random")$value)) == 276), "frequency"]`, respectively. The value 276 is the overall minimum (potentially the global minimum), while 421 is significantly worse. 

To compare the parameter vectors that led to these different values, we can use the `$closest_parameter()` method. From the saved optimization runs, it extracts the parameter vector corresponding to an optimum closest to `value`. We consider only results from the `nlm` optimizer here: 

```{r, closest parameters, eval = TRUE}
(mle <- mixture_ino$closest_parameter(value = 276, which_run = "random", which_optimizer = "nlm"))
mixture_ino$evaluate(at = as.vector(mle))
mle_run <- attr(mle, "run")
(bad <- mixture_ino$closest_parameter(value = 421, which_run = "random", which_optimizer = "nlm"))
mixture_ino$evaluate(at = as.vector(bad))
bad_run <- attr(bad, "run")
```

These two parameter vectors are saved as `mle` (this shall be our maximum likelihood estimate) and `bad` (this clearly is a bad estimate). Two attributes show the run id and the optimizer that led to these parameters.

To understand the values in terms of means, standard deviations, and mixing proportion (i.e., in the form $\boldsymbol{\theta} = (\mu_1, \mu_2, \sigma_1, \sigma_2, \lambda)$), they need transformation (see above):

```{r, transform parameter, eval = TRUE}
transform <- function(theta) c(theta[1:2], exp(theta[3:4]), plogis(theta[5]))
(mle <- transform(mle))
(bad <- transform(bad))
```

The two estimates `mle` and `bad` for $\boldsymbol{\theta}$ correspond to the following mixture densities:

```{r, estimated-mixtures, eval = TRUE}
mixture_density <- function (data, mu, sd, lambda) {
  lambda * dnorm(data, mu[1], sd[1]) + (1 - lambda) * dnorm(data, mu[2], sd[2])
}
ggplot(faithful, aes(x = eruptions)) + 
  geom_histogram(aes(y = after_stat(density)), bins = 30) + 
  labs(x = "eruption time (min)", colour = "parameter") +
  stat_function(
    fun = function(x) {
      mixture_density(x, mu = mle[1:2], sd = mle[3:4], lambda = mle[5])
    }, aes(color = "mle"), linewidth = 1
  ) +
  stat_function(
    fun = function(x) {
      mixture_density(x, mu = bad[1:2], sd = bad[3:4], lambda = bad[5])
    }, aes(color = "bad"), linewidth = 1
  )
```

The mixture defined by the `mle` parameter fits much better than `bad`, which practically estimates only a single class. However, the gradients at both points are close to zero, which explains why the `nlm` optimizer terminates at both points:

```{r, extract gradients, eval = TRUE}
mixture_ino$results(
  which_run = c(mle_run, bad_run), which_optimizer = "nlm", which_element = "gradient"
)
```

### Custom sampler for initial values

Depending on the application and the magnitude of the parameters to be estimated, initial values drawn from a standard normal distribution (which is the default behavior when calling `$optimize(initial = "random")`) may not be a good guess. We can, however, easily modify the distribution that is used to draw the initial values. For example, the next code snippet uses starting values drawn from a $\mathcal{N}(\mu = 2, \sigma = 0.5)$ distribution:

```{r, custom sampler}
sampler <- function() stats::rnorm(5, mean = 2, sd = 0.5)
mixture_ino$optimize(initial = sampler, runs = 100, label = "custom_sampler")
```

To obtain the first results of these optimization runs, we can use the `summary()` method. Note that setting `which_run = "custom_sampler"` allows filtering, which is the benefit of setting a `label` when calling `$optimize()`.

```{r, summary of custom sampler results, eval = TRUE}
summary(mixture_ino, which_run = "custom_sampler", digits = 2) |>
  head(n = 10)
```

Again we obtain different optima (even more than before). But in contrast, most of the runs here lead to the presumably global optimum of 276:

```{r, overview optima for custom sampler, eval = TRUE}
mixture_ino$optima(digits = 0, sort_by = "value", which_run = "custom_sampler")
```

### Educated guesses

Next we make "educated guesses" about starting values that are probably close to the global optimum. Based on the histogram above, the means of the two normal distributions may be somewhere around 2 and 4. We will use sets of starting values where the means are lower and larger than 2 and 4, respectively. For the variances, we set the starting values close to 1 (note that we use the log transformation here since we restrict the standard deviations to be positive by using `exp()` in the log-likelihood function). The starting value for the mixing proportion shall be around 0.5. This leads to the following 32 combinations of starting values:

```{r, fixed starting values, eval = TRUE}
mu_1 <- c(1.7, 2.3)
mu_2 <- c(4.3, 3.7)
sd_1 <- sd_2 <- c(log(0.8), log(1.2))
lambda <- c(qlogis(0.4), qlogis(0.6))
starting_values <- asplit(expand.grid(mu_1, mu_2, sd_1, sd_2, lambda), MARGIN = 1)
```

In the `$optimize()` method, instead of `initial = "random"`, we can set `initial` to a numeric vector of length `npar`, or, for convenience, to a `list` of such vectors, like `starting_values`:

```{r, optimization with educated guesses}
mixture_ino$optimize(initial = starting_values, label = "educated_guess")
```

These "educated guesses" lead to a way more stable optimization:

```{r, overview optima for educated guesses, eval = TRUE}
mixture_ino$optima(digits = 0, which_run = "educated_guess")
```

For comparison, we consider a set of implausible starting values...

```{r, bad guess}
mixture_ino$optimize(initial = rep(0, 5), label = "bad_guess")
```

... which lead to local optima:

```{r, bad guess summary, which_run = "random", eval = TRUE}
summary(mixture_ino, which_run = "bad_guess") 
```

### Standardizing the optimization problem

In some situations, it is possible to consider a standardized version of the optimization problem, which could potentially improve the performance of the numerical optimizer. In our example, we can standardize the data before running the optimization via the `$standardize()` method:

```{r, standardize data}
mixture_ino$standardize("data")
str(mixture_ino$get_argument("data"))
```

To optimize the likelihood using the standardized data set, we again use `$optmize()`, which by default uses random starting values. Below, we will compare these results with those obtained on the original optimization problem. 

```{r, optimization with standardized data}
mixture_ino$
  optimize(runs = 100, label = "data_standardized")$
  reset_argument("data")
```

The usage of `$reset_argument("data")` is important: to perform further optimization runs after having applied standardized initialization, we undo the standardization of the data and obtain the original data set. If we would not use `$reset_argument()`, all further optimization runs will be carried out on the standardized data set. 

### Reducing the optimization problem

In some situations, it is possible to first optimize a sub-problem and use those results as an initialization for the full optimization problem. For example, in the context of likelihood maximization, if the data set considered shows some complex structures or is very large, numerical optimization may become computationally costly. In such cases, it can be beneficial to initially consider a reduced data set. The following application of the `$reduce()` method transforms `"data"` by selecting a proportion of 30\% data points at random:

```{r, reduce data}
mixture_ino$reduce(argument_name = "data", how = "random", prop = 0.3, seed = 1)
str(mixture_ino$get_argument("data"))
```

Similar to the standardizing above, calling `$optimize()` now optimizes on the reduced data set:

```{r, optimization with reduced data}
mixture_ino$
  optimize(runs = 100, label = "data_subset")$
  reset_argument("data")$
  continue()
```

Again, we use `$reset_argument("data")` to obtain the original data set. The `$continue()` method now optimizes on the whole data set using the estimates obtained on the reduced data as initial values.

In addition to selecting sub samples at random (`how = "random"`), four other options exist via specifying the argument `how`:

- `"first"` selects the top data points,
- `"last"` selects the last data points,
- `"similar` selects similar data points based on k-means clustering,
- `"dissimilar"` is similar to `"similar"` but selects dissimilar data points.

See the other two vignettes for demonstrations of these options.

### Optimization times

The `$plot()` method provides an overview of the optimization times. Setting `by = "label"` allows for comparison across initialization strategies, setting `relative = TRUE` plots relative differences to the median of the top boxplot:

```{r, plot-by-label, eval = TRUE}
mixture_ino$plot(by = "label", relative = TRUE, xlim = c(-1, 3))
```

Setting `by = "optimizer"` allows comparison across optimizers:

```{r, plot-by-optimizer, eval = TRUE}
mixture_ino$plot(by = "optimizer", relative = FALSE, xlim = c(0, 0.05))
```

### The global optimum

The best optimization result can be extracted via:

```{r, extract best value and parameter, eval = TRUE}
mixture_ino$best_value()
mixture_ino$best_parameter()
```

The best function value of `r round(mixture_ino$best_value(), 2)` is unique:

```{r, best optimum, eval = TRUE}
head(mixture_ino$optima(digits = 0, sort_by = "value"))
```

Furthermore, it does not produce a two-class mixture, since one class variance is practically zero, see the output of `mixture_ino$best_parameter()`. We could therefore delete this result from our `Nop` object:

```{r, delete optimum, eval = TRUE}
mixture_ino$clear(which_run = attr(mixture_ino$best_value(), "run"))
```

The final `Nop` object then looks as follows:

```{r, print final mixture_ino object, eval = TRUE}
print(mixture_ino)
```

## References
