---
title: "Specify numerical optimizer"
author: "Lennart Oelschläger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specify numerical optimizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how to specify the numerical optimizer for which we seek to find good initial points. The vignette was build using R `r paste(R.Version()[6:7], collapse = ".")` with the ino `r utils::packageVersion("ino")` package.

```{r setup}
library(ino)
```

## The `set_optimizer()` function

The `set_optimizer()` function must be used to specify the numerical optimizer. It takes the following arguments:

```{r, args set_optimizer}
args(set_optimizer)
```

The input `x` must be either `NULL` or an object of class `ino`. In the former case, an `ino` object is initialized by the `set_optimizer()` function, in the latter case the `ino` object is updated.

The input `optimizer` must be either a character [specifying a pre-defined optimizer](#pre-defined-optimizer) or a function for [setting a custom optimizer](#custom-optimizer). 

The ellipsis input `...` takes additional arguments to be passed to the optimizer.

### Pre-defined optimizer

Specify the `optimizer` argument to one of the following characters to set currently pre-defined optimizerin in ino^[Extending this list is on our agenda.]:

- `"nlm"` for the Newton-type algorithm [nlm](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/nlm.html)

For example, calling

```{r, set_optimizer example}
set_optimizer(optimizer = "nlm", iterlim = 1000)
```

returns an `ino` object where the nlm optimizer with iteration limit 1000 is specified.

### Custom optimizer

The `optimizer` argument also can be a function to set a custom optimizer (or any optimizer not covered above). This functionality will be implemented in the future.

## Comparing optimizer

In the future, `ino` will provide the functionality of comparing the initialization effect among different optimizer. In order to do so, `set_optimizer()` must be used to set multiple optimizer.
