---
title: "Specify function"
author: "Lennart Oelschläger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specify function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how to specify the function that we seek to optimize numerically. The vignette was build using R `r paste(R.Version()[6:7], collapse = ".")` with the ino `r utils::packageVersion("ino")` package.

```{r setup}
library(ino)
```

## The `set_f()` function

The `set_f()` function must be used to specify the function. It takes the following arguments:

```{r, args set_f}
args(set_f)
```

The input `x` must be either `NULL` or an object of class `ino`. In the former case, an `ino` object is initialized by the `set_f()` function, in the latter case the suplied `ino` object is updated.

The input `f` must be an object of class `function` and specifies the function to be optimized.

The input `npar` is the number of parameters in `f`.

The ellipsis argument `...` can be used to specify additional arguments in `f`.

## Example

```{r, example set_f}
himmelblau <- function(x, c, d) (x[1]^2 + x[2] + c)^2 + (x[1] + x[2]^2 + d)^2
set_f(f = himmelblau, npar = 2, c = -11, d = -7)
```
